# Results

Our main hypothesis was that there is likely some demographic that is facing more hardship due to medication price inflation in counties with these demographics within the US. Can we use the demographics of a specific county along with the now know medication prices within said county to develop a model to predict the medication prices? If yes, we could in the short term advise people with specific medical condistions to move to counties with demographics that provide more favorable medication prices and long term influence the industry to fix such inequities. 

 An early concern in the development of this project was whether there would even be any price variation based on location.  A common misconception is that Medicare drug prices are the same across the US, given that Medicare is a public national medical insurance. The results were surprising. Drug prices do, in fact, vary significantly regionally, sometimes at the state level and sometimes at the county level. Semaglutide is one drug that stands out at the state level as shown in @fig-map-semaglutide. New York state has the lowest median 30-day supply price compared to every other stat in the country.

```{r}
#| label: fig-map1
#| fig-cap: "Median Medicare Price for 30-day Supply of Semaglutide by US County"
#| warning: false

chosen_drug = "Semaglutide"

# Load libraries
library(tigris)
library(sf)
library(ggplot2)
library(dplyr)

summary_table <- readRDS("data/map_data.rds")

# Use sf-compatible county shapefile (from tigris)
options(tigris_use_cache = TRUE)
counties_sf <- tigris::counties(cb = TRUE, class = "sf")

# Change county_code to FIPS style
# Choose drug of interest
summary_table2 <- summary_table %>%
  mutate(COUNTY_CODE = sprintf("%05s", COUNTY_CODE)) %>%
  filter(drug_name_clean == chosen_drug)

# Join median_cost to county map
counties_joined <- counties_sf %>%
  left_join(summary_table2, by = c("NAME" = "COUNTY", "STUSPS" = "STATE"))

# Define bounding box for continental US
bbox <- st_bbox(c(xmin = -125, xmax = -66, ymin = 23, ymax = 50), crs = st_crs(counties_joined))

# Crop the map
counties_cropped <- st_crop(counties_joined, bbox)

# Plot it
ggplot(counties_cropped) +
  geom_sf(aes(fill = median_price)) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    title = paste0("Median Medicare Price for 30-day Supply of ",  chosen_drug, " by US County "),
    fill = "Median Cost"
  ) +
  theme_minimal(base_size = 14)
```

Liraglutide is unique in that it has a clear state-wide price impact like in New Mexico and Minnesota, but a county level price delta in some counties scattered across other states such as Nebraska, Oregon, Washington, Tennessee, Missouri, and Virginia. {@fig-map2}

```{r}
#| label: fig-map2
#| fig-cap: "Median Medicare Price for 30-day Supply of Liraglutide by US County"
#| warning: false

chosen_drug = "Liraglutide"

# Load libraries
library(tigris)
library(sf)
library(ggplot2)
library(dplyr)

summary_table <- readRDS("data/map_data.rds")

# Use sf-compatible county shapefile (from tigris)
options(tigris_use_cache = TRUE)
counties_sf <- tigris::counties(cb = TRUE, class = "sf")

# Change county_code to FIPS style
# Choose drug of interest
summary_table2 <- summary_table %>%
  mutate(COUNTY_CODE = sprintf("%05s", COUNTY_CODE)) %>%
  filter(drug_name_clean == chosen_drug)

# Join median_cost to county map
counties_joined <- counties_sf %>%
  left_join(summary_table2, by = c("NAME" = "COUNTY", "STUSPS" = "STATE"))

# Define bounding box for continental US
bbox <- st_bbox(c(xmin = -125, xmax = -66, ymin = 23, ymax = 50), crs = st_crs(counties_joined))

# Crop the map
counties_cropped <- st_crop(counties_joined, bbox)

# Plot it
ggplot(counties_cropped) +
  geom_sf(aes(fill = median_price)) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    title = paste0("Median Medicare Price for 30-day Supply of ",  chosen_drug, " by US County "),
    fill = "Median Cost"
  ) +
  theme_minimal(base_size = 14)
```

Our team was surprised by what we observed once the median prices were applied to map plots.  There are clearly patterns here, but they vary geographically and the pattern is not the same with different drugs. To see maps for many more drugs, see this [shiny app](https://gpc1.shinyapps.io/shiny_maps_drug_median_price/). 

Given the pattern for medication prices across the maps, we started a deep dive into the demographics. Can we find any demographic data that can help predict thes prices? Our first dive was into the population. Do counties with a larger population have more Medicare plan options and this competition leads to lower drug prices? A scatter plot of the number of Medicare Plans available vs the population per county is shown below with a linear fit to show a trend line. {@fig-plans-vs-population} While there is a slight trend, it is a not a great fit. 

```{r}
#| label: fig-plans-vs-population
#| fig-cap: "Correlation between the number of Medicare Plan Options and Total Population by US County in 23Q4"
#| warning: false


library(tidyverse)
library(ggplot2)

summary_table = readRDS("data/summary_table_23Q4.rds")

summary_table2 <- summary_table %>%
  filter(!is.na(total_popE))

# Plot population vs
ggplot(summary_table2, aes(x = num_contract_plans, y = total_popE)) +
  geom_point(alpha = 0.6, color = "gray") +
  scale_y_log10(labels = scales::comma) + 
  geom_smooth(method = "lm", color = "black", se = FALSE, linetype = "dashed") +
  
  labs(
    title = "Number of Medicare Plan Options vs Total Population by US County in 23Q4",
    x = "Number of Medicare Plan Options",
    y = "Total Population (log scale)"
  ) +
  theme_minimal(base_size = 14)
```

While there is a slight trend, it is not a great fit overall. The model is statistically significant (p-vaule < 0.05) and shows a moderate positive relationship (slope = 0.0195) between number of Medicare plans and population. But it's not a strong fit, two-thirds of the variation in population is left unexplained (adjusted R-squared = 0.3444), so we don’t wnat over-interpret this model as being predictive.
 
From here, we jumped into creating a model across all our demographics (age, education, gender, median income, and percent below poverty line). 

***ADD MACHINE LEARNING STUFF***

This regional variation underscores the importance of localized analysis. It also raises important questions about equity and access—particularly for beneficiaries in high-cost areas who may face greater financial burdens for the same treatments available at lower costs elsewhere. These disparities suggest that systemic inefficiencies or market imbalances may be at play, warranting further investigation and policy attention.

This regional variation underscores the importance of localized analysis. It also raises important questions about equity and access—particularly for beneficiaries in high-cost areas who may face greater financial burdens for the same treatments available at lower costs elsewhere. These disparities suggest that systemic inefficiencies or market imbalances may be at play, warranting further investigation and policy attention.


