# Results

Our main hypothesis was that there is likely some demographic that is facing more hardship due to medication price inflation in counties with these demographics within the US. Can we use the demographics of a specific county, along with the median medication prices within said county, to develop a model to predict the medication prices? If yes, we could in the short term advise people with specific medical conditions to move to counties with demographics that provide more favorable medication prices and long-term influence the industry to fix such inequities. 

## Statistical Testing - Phase 1

To consider the impact to different demographics by varying prices for the selected drugs, our team constructed a series of hypotheses.  The first consideration was towards race.  The majority of U.S. counties contain percentages which heavily tilt towards individuals who identify as white.  In fact, when evaluating the percentages across all counties throughout the US, counties are 95% white at the median and 93% at the mean.  For our first test, we elected to evaluate counties that were at least 95% white and those below that threshold.

Before proceeding, we had to address assumptions which would be critical to determining which tests could actually apply.  We did not initially feel confident applying the assumption of normality to our data.  The spread of the median and mean price values across counties and within various medication groups did visually appear normal.  However, after applying the Shapiro-Wilk test, Shapiro p-values were all significantly below 0.05 which indicated a lack of normalcy.  Accordingly our team chose a non-parametric test to evaluate our hypotheses.  The Anderson-Darling test was explored as an alternative to the Shapiro-Wilk test, but the results were similar, with high AD statistic values, suggesting a lack of normality. 

The Wilcoxon rank-sum (Mann-Whitney U) test was selected to evaluate two independent groups, specified by the hypothesis threshold.  We considered the observations to be independent of each other since the census is completed at the county level, with no counties overlapping.

Ultimately, the Mann-Whitney U Test is less powerful than parametric tests when assumptions are met.  It also lacks confidence intervals and is only comparing the ranks and not the actual values.  The benefit of taking this approach for our data is that fewer overall assumptions are necessary.  Since our team was condensing a number of variables, this approach was preferred to starting with an excessive number of assumptions.

### Hypothesis 1

*Null Hypothesis:* The distribution of medication costs in counties with >=95% white population is not statistically different from the distribution in counties with <95% white population.

*Alternative Hypothesis:* The distribution of medication costs in counties with >= 95% white population is statistically different from that in counties with <95% white population.

*Assumptions:* 

1. – Independent groups
2. – Ordinal or continuous data
3. – Same Shape of Distribution (when comparing medians)
4. – Random Sampling
5. – Independence within each group

Accordingly, our team elected to compare mean prices, which notably exposes our analysis to outlier pricing.  The assumption independence within each group is stretched, considering the manner in which the data was procured, which we have discussed in great detail.

| Drug Name               | p-value        |
|------------------------|----------------|
| Adalimumab             | 3.37e-21        |
| Apixaban               | 5.46e-21        |
| Insulin Aspart         | 6.31e-18        |
| Semaglutide            | 2.53e-13        |
| Paliperidone Palmitate | 6.19e-13        |
| Rivaroxaban            | 4.67e-12        |
| Liraglutide            | 1.44e-10        |
| Empagliflozin          | 4.23e-04        |
| Tiotropium Bromide     | 1.16e-02        |
| Dulaglutide            | 1.30e-02        |
| Lenalidomide           | 6.55e-02        |
| Etanercept             | 7.12e-02        |
| Ustekinumab            | 7.29e-02        |
| Ibrutinib              | 4.93e-01        |

We can reasonably conclude from this that in 10 of our 14 focus medications, there is a statistically significant difference in mean prices and reject the null hypothesis for these medications.  To dive deeper, our team constructed density plots to compare the direction of this difference.  As observed in {@fig-h1}, where we evaluate two of these medications.  We can clearly see that the direction shifts at different price ranks and depends on the medication.

```{r}
#| label: fig-h1
#| fig-cap: "Hypothesis 1 - Comparing GLP-1 receptor agonists, primarily used for type 2 diabetes management"
#| warning: false
#| echo: false

library(tidyverse)

h_df_clean = readRDS("data/h_df_clean.rds")

h_df_clean%>%
  filter(drug_name_clean %in% c("Dulaglutide", "Semaglutide"))%>%
  group_by(drug_name_clean) %>%
  filter(
    mean_price > quantile(mean_price, 0.25) - 1.5 * IQR(mean_price) &
      mean_price < quantile(mean_price, 0.75) + 1.5 * IQR(mean_price)
  )%>%
  ungroup()%>%
  ggplot(aes(x = mean_price, fill = white_95)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("blue", "orange"),
                    labels = c("<95% White", "≥95% White")) +
  labs(
    title = "Density Plots of Mean Drug Prices by County Demographics",
    x = "Mean Price",
    y = "Density",
    fill = "County Group",
    caption = "*Outliers removed"
  ) +
  scale_x_continuous(labels = scales::label_dollar(accuracy=1))+
  facet_grid(.~ drug_name_clean, scales = "free") +
  theme_minimal()+
  theme(panel.spacing = unit(1.5, "lines")) 


```

Drilling down again, with these two medications in focus, actual values are compared:

| Drug Name   | >=95% White | Mean of Mean Price | SD Mean Price |
|-------------|----------|----------------|----------------|
| Dulaglutide | FALSE    | 444.00          | 3.74           |
| Dulaglutide | TRUE     | 444.00          | 3.71           |
| Semaglutide | FALSE    | 168.00          | 2.58           |
| Semaglutide | TRUE     | 167.00          | 1.84           |

The findings are interesting but have issues.  In the plot, we can clearly see that advantages are given to certain groups at certain prices.  The mean of the mean prices, as well as the standard deviations show practically no difference for Dulaglutide and only some difference for Semaglutide.  Unfortunately, this does not provide much clarity regarding any particular advantage afforded to counties where more than 95% of the population identifies as white.  The findings are hardly definitive and lack confidence intervals.  This does demonstrate the limitations of using the Mann-Whitney U Test.

### Hypothesis 2

Using the same assumptions, another consideration involves fully removing the racial aspect and instead considering the percentage of the population that is below the poverty level.  Within the census data, the median is 13% below the poverty level, across all counties and 14% at the mean.  We establish a new hypothesis.

*Null Hypothesis:* The distribution of medication costs in counties with >=13% below poverty population is not statistically different from the distribution in counties with <13% below poverty population.

*Alternative Hypothesis:* The distribution of medication costs in counties with >= 13% below poverty population is statistically different from that in counties with <13% below poverty population.

| Drug Name               | p-value        |
|------------------------|----------------|
| Semaglutide            | 1.43e-20        |
| Dulaglutide            | 2.49e-16        |
| Ibrutinib              | 4.41e-13        |
| Insulin Aspart         | 2.34e-10        |
| Adalimumab             | 3.46e-10        |
| Etanercept             | 2.40e-08        |
| Ustekinumab            | 8.88e-08        |
| Apixaban               | 1.90e-06        |
| Empagliflozin          | 6.70e-06        |
| Liraglutide            | 1.12e-03        |
| Lenalidomide           | 1.43e-03        |
| Paliperidone Palmitate | 2.70e-02        |
| Tiotropium Bromide     | 8.55e-02        |
| Rivaroxaban            | 1.54e-01        |

Again, we find that there are statistically significant outcomes for many of the selected medications.  Twelve of the fourteen medications yield p-values below 0.05 suggesting that the null hypothesis be rejected.

In {@fig-h2} we look closer at six medications, to expand the investigation into these differences and the direction of any potential advantage.

```{r}
#| label: fig-h2
#| fig-cap: "Hypothesis 2 - Comparing six medications which were indicated for statistically significant differences in price for two poverty groups"
#| warning: false
#| echo: false

library(tidyverse)

h_df_clean = readRDS("data/h_df_clean.rds")

h_df_clean%>%
  filter(drug_name_clean %in% c("Semaglutide", "Dulaglutide"))%>%
  group_by(drug_name_clean) %>%
  filter(
    mean_price > quantile(mean_price, 0.25) - 1.5 * IQR(mean_price) &
      mean_price < quantile(mean_price, 0.75) + 1.5 * IQR(mean_price)
  )%>%
  ungroup()%>%
  ggplot(aes(x = mean_price, fill = poverty_13)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("blue", "orange"),
                    labels = c("<13% Below Poverty", "≥13% Below Poverty")) +
  labs(
    title = "Density Plots of Mean Drug Prices by County Demographics",
    x = "Mean Price",
    y = "Density",
    fill = "County Group",
    caption = "*Outliers removed"
  ) +
  scale_x_continuous(labels = scales::label_dollar())+
  facet_wrap(~ drug_name_clean, scales = "free") +
  theme_minimal()


```

Similar to hypothesis 1, we see that the price differences are contextual to the medication and price rank position.  While the null hypothesis is rejected, the challenge of approaching multiple medications is clearly demonstrated.  For both hypotheses, by rejecting the null, this may result in a type 1 error, or false positive.  There are clearly price differences relative to certain demographic groups.  Whether those differences are being driven by the demographics remains unclear.

## Population vs Number of Available Medicare Plans

Given the lack of clear patterns for medication prices across the maps, we turned to a deep dive into the demographics. Our first dive was into the population. Do counties with a larger population have more Medicare plan options, and does this competition lead to lower drug prices? A scatter plot of the number of Medicare Plans available vs the population per county is shown below with a linear fit to show a trend line. {@fig-plans-vs-population} 


```{r}
#| label: fig-plans-vs-population
#| fig-cap: "Correlation between the number of Medicare Plan Options and Total Population by US County in 23Q4"
#| warning: false
#| echo: false


library(tidyverse)
library(ggplot2)

summary_table = readRDS("data/summary_table_23Q4.rds")

summary_table2 <- summary_table %>%
  filter(!is.na(total_popE))

# Plot population vs
ggplot(summary_table2, aes(x = num_contract_plans, y = total_popE)) +
  geom_point(alpha = 0.6, color = "gray") +
  scale_y_log10(labels = scales::comma) + 
  geom_smooth(method = "lm", color = "black", se = FALSE, linetype = "dashed") +
  
  labs(
    title = "Number of Medicare Plan Options vs Total Population by US County in 23Q4",
    x = "Number of Medicare Plan Options",
    y = "Total Population (log scale)"
  ) +
  theme_minimal(base_size = 10) + 
  theme(
    plot.title.position = "plot",       # aligns title with plot area
    plot.title = element_text(hjust = 0) # hjust = 0 means left-align
  )

```

While there is a slight trend, it is not a great fit overall. The model is statistically significant (p-value < 0.05) and shows a moderate positive relationship (slope = 0.0195, or 1 more medicare plan with a 1.95% increase in the population) between the number of Medicare plans and population. But it's not a strong fit. Two-thirds of the variation in population is left unexplained (adjusted R-squared = 0.3444), so we don’t want to over-interpret this model as being predictive. Given the limited correlation between the number of available plans and population, we don’t think this demographic measure is having much of an impact on medication prices. 
 
From here, we jumped into creating a model across all our demographics (age, education, gender, and racial breakdowns, as well as median income, and percent below the poverty line). 



## Machine Learning

### Check for Gaussian Distributions

Before jumping into any modeling, one must check the distribution of the variables. For this particular dataset, if all the medications are looked at simultaneously, one needs to take the log of the median prices to have the histogram show a more Gaussian distribution. (See @fig-hist-price-all-drugs). This is due to the fact that the median drug price varies greatly depending on the actual drug being considered. 

```{r}
#| label: fig-hist-price-all-drugs
#| fig-cap: "Histogram of Median Drug Prices and Log-Transformed Median Drug Prices across All Drugs"
#| warning: false
#| fig-height: 4
#| echo: false


summary_table2 = readRDS("data/summary_table2.rds")

# Set up a 1-row, 2-column layout
par(mfrow = c(1, 2))
# Histogram of raw median_price
hist(summary_table2$median_price, 
     breaks = 50, 
     main = "Median Price - All Drugs", 
     xlab = "Price",
     col = "darkgray")
# Histogram of log-transformed median_price
hist(log(summary_table2$median_price), 
     breaks = 50, 
     main = "Log(Median Price) - All Drugs", 
     xlab = "log(Price)",
     col = "lightgray")
# Reset plotting layout
par(mfrow = c(1, 1))
```

If the medications are reviewed independently (@fig-hist-price-per-drug), then we do not need to take the log of the medication prices. The distribution shape doesn’t change with or without the log of the price. Our plan is to look at each drug independently due to the lack of patterns in the correlation heatmaps earlier (@fig-by-drug-correlations).

```{r}
#| label: fig-hist-price-per-drug
#| fig-cap: "Histogram of Median Drug Prices by Drug"
#| warning: false
#| echo: false
#| fig-height: 8

summary_table2 = readRDS("data/summary_table2.rds") %>% filter(!is.na(drug_name_clean))

library(dplyr)
library(ggplot2)
# Plot faceted histograms
ggplot(summary_table2, aes(x = median_price)) +
  geom_histogram(bins = 50, fill = "darkgray") +
  facet_wrap(~ drug_name_clean, scales = "free", ncol = 3) +
  labs(
    title = "Distribution of Median Price by Drug",
    x = "Median Price",
    y = "Count"
  ) +
  theme_minimal()

```

### Model Selection
In order to model the relationship between medication prices and demographic features, we employed a variety of machine learning algorithms, each with unique strengths and assumptions. Below is a brief overview of the algorithms explored:

**Linear Model (LM):**
Linear regression is a fundamental technique used to model the relationship between a continuous outcome variable and one or more predictors, assuming a linear relationship. It is interpretable and computationally efficient but may underperform when the underlying patterns are nonlinear or complex.


**Elastic Net (ENET):**
Elastic Net is a regularized regression method that combines both L1 (Lasso) and L2 (Ridge) penalties. It is particularly effective when dealing with multicollinearity or when performing variable selection in high-dimensional datasets. Elastic Net encourages sparsity while maintaining group stability among correlated features.


**Gradient Boosting Machine (GBM):**
GBM is an ensemble technique that builds models sequentially, with each new model correcting the errors of the previous one. It is capable of capturing complex nonlinear relationships and is known for its high predictive performance, although it can be prone to overfitting if not properly tuned.


**K-Nearest Neighbors (KNN):**
KNN is a non-parametric algorithm that classifies or predicts outcomes based on the average value (or majority class) of the k closest observations in the feature space. It is simple and intuitive but can struggle with high-dimensional data and is sensitive to the choice of distance metric and value of k.


**Random Forest (RF):**
Random Forest is an ensemble method that constructs a large number of decision trees during training and outputs the average prediction (for regression) or majority vote (for classification). It offers good performance with low risk of overfitting, handles missing data well, and provides insights into feature importance.


**Support Vector Machine (SVM):**
SVM is a powerful classifier and regressor that aims to find the optimal hyperplane that maximally separates classes or fits data with minimal error (via support vectors). It works well in high-dimensional spaces and with complex boundaries but can be computationally intensive and sensitive to kernel and parameter choices.


**Extreme Gradient Boosting (XGBoost):**
XGBoost is an optimized and scalable implementation of gradient boosting. It includes advanced features such as regularization, parallel processing, and handling of missing values. XGBoost typically outperforms many other algorithms in structured data problems and is widely used in machine learning competitions and real-world applications.


### Model Performance

Looping over every drug, we trained all the previously described models. A heatmap of the resulting model performance is provided below in @fig-heatmap-models-performance. Note that all models for Ustekinumab have outlier RMSE and MAE results and are thus gray to make the difference between the other drug results more differentiated visually. For all other drugs, the random forest model has the lowest RMSE and MAE values. Fortunately, the random forest model also provides insight into feature importance. 


```{r}
#| label: fig-heatmap-models-performance
#| fig-cap: "Model Performance Heatmap by Drug"
#| warning: false
#| echo: false
#| fig-height: 6

library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)

# # Load and prepare data
# model_metrics_df <- readRDS("data/model_metrics_summary.rds")
# 
# metrics_long <- model_metrics_df %>%
#   pivot_longer(cols = c(RMSE, MAE), names_to = "Metric", values_to = "Value") %>%
#   # Normalize Value to [0,1] to assess brightness
#   mutate(fill_norm = rescale(Value),
#          text_color = ifelse(fill_norm > 0.05, "white", "black"))  # text is black if background is light
# 
# ggplot(metrics_long, aes(x = model, y = drug, fill = Value)) +
#   geom_tile(color = "white") +
#   geom_text(aes(label = signif(Value, 2), color = text_color), size = 3, angle = 45, show.legend = FALSE) +
#   scale_color_identity() +  # use colors as-is from the column
#   scale_fill_viridis_c(option = "magma", direction = -1, name = "Metric Value",     
#                        limits = c(min(metrics_long$Value, na.rm = TRUE), 
#                                   quantile(metrics_long$Value, 0.90, na.rm = TRUE))) +
#   facet_wrap(~ Metric, ncol = 2) +
#   labs(
#     title = "Model Performance by Drug (Dynamic Text Contrast)",
#     x = NULL, #"Model",
#     y = NULL #"Drug"
#   ) +
#   theme_minimal(base_size = 12) +
#   theme(
#     axis.text.x = element_text(angle = 45, hjust = 1),
#     plot.title = element_text(hjust = 0),
#     plot.title.position = "plot"
#   )

all_metrics = readRDS("data/all_metrics_untuned_models.rds")

train_metrics <- all_metrics %>% 
  filter(Dataset == "Train") %>% 
  mutate(fill_norm = rescale(Rsquared), 
         text_color = ifelse(fill_norm > 0.25, "white", "black"))  # text is black if background is light
test_metrics  <- all_metrics %>% 
  filter(Dataset == "Test") %>% 
  mutate(fill_norm = rescale(Rsquared), 
         text_color = ifelse(fill_norm > 0.5, "white", "black"))  # text is black if background is light

# Shared fill scale
Rsq_limits <- c(min(all_metrics$Rsquared, na.rm = TRUE),
                quantile(all_metrics$Rsquared, 1, na.rm = TRUE))

# Train Rsquared heatmap
p1 <- ggplot(train_metrics, aes(x = Model, y = Drug, fill = Rsquared)) +
  geom_tile(color = "white") +
  geom_text(aes(label = signif(Rsquared, 2),
                color = text_color),
            size = 3, angle = 45, 
            show.legend = FALSE) +
  scale_color_identity() +  # use colors as-is from the column
  scale_fill_viridis_c(option = "magma", direction = -1, name = "R^2", limits = Rsq_limits) + 
  labs(title = "Train Rsquared", x = "Model", y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Test Rsquared heatmap
p2 <- ggplot(test_metrics, aes(x = Model, y = Drug, fill = Rsquared)) +
  geom_tile(color = "white") +
  geom_text(aes(label = signif(Rsquared, 2),
                color = text_color),
            size = 3, angle = 45, 
            show.legend = FALSE) +
  scale_color_identity() +  # use colors as-is from the column
  scale_fill_viridis_c(option = "magma", direction = -1, name = "R^2", limits = Rsq_limits) +
  labs(title = "Test Rsquared", x = "Model", y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
# Combine
(p1 + p2) + plot_layout(guides = "collect") & theme(legend.position = "right")

```


### Feature Importance
 The most important features for the random forest models across all drugs are shown in the heatmap below (@fig-heatmap-rf-variable-importance). The most obvious pattern here is that the num_contract_plans (aka the quantity of available Medicare plans) is the most important feature for 10 out of the 14 top drugs. We already looked at the connection between total_population and the quantity of available plans in a county previously, and only saw a mild correlation. For the other four drugs, the top features were percent of residences with an associate's degree as their maximum education level, the median household income, the percentage of the population that is black, and the percentage of the population where the highest level of education is a master’s degree.


``` {r}
#| label: fig-heatmap-rf-variable-importance
#| fig-cap: "Random Forest Variable Importance Heatmap by Drug"
#| warning: false
#| echo: false
#| fig-height: 6

library(ggplot2)
top_rf_vars2 <- readRDS("data/top_rf_vars2.rds")

ggplot(top_rf_vars2, aes(x = Variable, y = Drug, fill = Overall)) +
  geom_tile(color = "white") +
  geom_text(aes(label = signif(Overall, 2), color = label_color), size = 3) +
  scale_color_identity() +  # use colors directly from label_color
  scale_fill_viridis_c(option = "magma", name = "Importance") +
  coord_flip() +
  labs(
    title = "Random Forest Variable Importance Heatmap by Drug",
    x = NULL, #"Variable",
    y = NULL #"Drug"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)
  )

```

### Machine Learning (ML) Conclusions

The application of diverse machine learning algorithms provided valuable insights into the complex relationships between demographic factors and medication prices across U.S. counties. Despite the initial hypothesis that demographic variables might strongly predict drug pricing disparities, our results indicate that the quantity of Medicare plans available within a county is the most consistent and influential predictor across the majority of the top-spending medications. Other demographic features, such as education level, income, and racial composition, showed variable importance depending on the specific drug but did not emerge as universally strong predictors.

The superior performance of random forest modelling highlights the nonlinear and multifaceted nature of drug pricing patterns, which simpler linear models struggled to capture effectively. This complexity underscores the challenge of modeling medication prices purely based on demographics and points to the potential influence of additional factors not captured in our current dataset.

Overall, these findings suggest that while demographic characteristics contribute to medication price variation, market dynamics (reflected by Medicare plan availability) play a pivotal role. Future work may benefit from incorporating additional economic, regulatory, and supply-chain data to better explain and predict geographic disparities in medication costs. This knowledge could ultimately inform policies aimed at improving affordability and equity in drug pricing.














